---
title: "Assignment Week 05"
author: "Bohan Zhang"
date: "2025-03-24"
output: html_document
---

Here we use the H3K4me3 and H3K27me3 peaks from mouse embryonic stem cells (ESC).

```{r setup, include=FALSE}
library(rtracklayer)
library(GenomicRanges)
setwd("/mnt/gapp/work/bohan/bioinfo/Week05/")
```

## Get the data from last week

```{r}
H3K4me3 <- import("/mnt/gapp/work/bohan/bioinfo/Week04/H3K4me3_peaks.bed.gz", format = "narrowPeak")
H3K27me3 <- import("/mnt/gapp/work/bohan/bioinfo/Week04/H3K27me3_peaks.bed.gz", format = "narrowPeak")
head(H3K4me3)
```

## Identify bivalent domains

Here the "bivalent domains" means regions that have both H3K4me3 and H3K27me3 mark.

```{r}
bivalent <- findOverlaps(H3K4me3, H3K27me3, type = "any")
bivalent_domains <- pintersect(H3K4me3[from(bivalent)],
                              H3K27me3[to(bivalent)])
bivalent_domains
```

Here we identified 4191 bivalent domains.

## Upon differentiation

### Download files for differentiation cells

Here we choose B cells as differentiated cell type.

```{r}
download.file("https://www.encodeproject.org/files/ENCFF496DOC/@@download/ENCFF496DOC.bed.gz", 
              "H3K4me3_Bcell.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF931RRP/@@download/ENCFF931RRP.bed.gz", 
              "H3K27me3_Bcell.bed.gz")
H3K4me3_Bcell <- import("H3K4me3_Bcell.bed.gz", format = "narrowPeak")
H3K27me3_Bcell <- import("H3K27me3_Bcell.bed.gz", format = "narrowPeak")
```
### Investigate the overlaps

Now we investigate how many of the ESC bivalent domains are overlapping either mark or their combination in B cells:

```{r}
# Only overlap with H3K4me3:
H3K4me3_single_mark_overlap <- findOverlaps(bivalent_domains, H3K4me3_Bcell, type = "any")
length(H3K4me3_single_mark_overlap)

#Only overlap with H3K27me3:
H3K27me3_single_mark_overlap <- findOverlaps(bivalent_domains, H3K27me3_Bcell, type = "any")
length(H3K27me3_single_mark_overlap)

#Overlap with both marks:
com <- findOverlaps(H3K4me3_Bcell, H3K27me3_Bcell, type = "any")
com_domains <- pintersect(H3K4me3_Bcell[from(com)],
                          H3K27me3_Bcell[to(com)])
head(com_domains)
overlap_combination <- findOverlaps(bivalent_domains, com_domains, type = "any")
length(overlap_combination)
```

### Final report

```{r}
cat("number of ESC bivalent domains overlaping with H3K4me3: ", length(H3K4me3_single_mark_overlap),
    "\nnumber of ESC bivalent domains overlaping with H3K27me3: ", length(H3K27me3_single_mark_overlap),
    "\nnumber of ESC bivalent domains overlaping with the combination of the two marks: ",  length(overlap_combination))
```


