---
title: "Assignment Week 12"
author: "Bohan Zhang"
date: "2025-05-26"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(AnnotationHub)
  library(ensembldb)
  library(InteractionSet)
  library(ggplot2)
})
setwd("/mnt/gapp/work/bohan/bioinfo/Week12")
```

## Data Preparation 

We chose p300 as in the practical during the course:

```{r}
# p300 peaks in A549 cells:
peaks <- rtracklayer::import("p300.narrowPeak.gz")
seqlevelsStyle(peaks) <- "ensembl"
head(peaks)
```

Granges file from Salviato et al.(2021) for peaks with predicted distal targets

```{r}
enhancerTar <- readRDS("hg38.SalviatoDistalEnhancerTargets.GR.rds")
seqlevelsStyle(enhancerTar) <- "ensembl"
head(enhancerTar)
```

## Subset peaks and label distal targets of peaks

```{r}
# find the peaks that have distal target
o <- findOverlaps(peaks, enhancerTar)
head(o)
# label the peaks with targets:
mcols(peaks)[from(o),"target"] <- enhancerTar[to(o)]$target
mcols(peaks)$target <- CharacterList(mcols(peaks)$target)
# Only keep the peaks that are overlapped with the enhancers:
peaks <- peaks[from(o), ]
# look at the output
peaks
```

## Find the nearest TSS for each peak

```{r}
# Firstly load the reference genome
ah <- AnnotationHub()
ensdb <- ah[["AH95744"]] # human ensembl 104 annotation
genes <- genes(ensdb)
# get all the TSS from the reference genome
tss <- resize(genes, width = 1, fix = "start")

# find the nearest TSS for each peak
nearest_tss <- tss[nearest(peaks, tss)]
peaks$nearest_tss_gene <- nearest_tss$gene_name
peaks$tss_start <- start(nearest_tss)
# Some of them doesn't have symbol ID. Remove them:
peaks <- peaks[peaks$nearest_tss_gene!="",]
head(peaks)
```

Some rows look exactly the same. We should remove them:

```{r}
peaks <- peaks[!duplicated(peaks)]
head(peaks)
```

## Split the peaks into upstream or downstream of the predicted target:

```{r}
relative_pos <- start(peaks) - peaks$tss_start
peaks$position <- ifelse(relative_pos < 0, "upstream", "downstream")
head(peaks)
```

## Calculate the proportion that the predicted target is the closest gene:

```{r}
same_gene <- peaks$target == peaks$nearest_tss_gene
upstream_prop <- mean(sapply(same_gene[peaks$position == "upstream"], any))
downstream_prop <- mean(sapply(same_gene[peaks$position == "downstream"], any))

print(paste0("The proportion of upstream set: ", sprintf("%.2f%%", upstream_prop * 100), 
             ". The proportion of downstream set: ", sprintf("%.2f%%", downstream_prop * 100)))

```

## Plot the distribution of the distances between peaks and TSS of the predicted target genes

```{r}
# Calculate the absolute value of the distance
distances <- abs(start(peaks) - peaks$tss_start)
peaks$distance <- distances

# Visualization
ggplot(as.data.frame(peaks), aes(x=distance, fill=position)) +
  geom_histogram(bins=50, alpha=0.6, position="identity") +
  scale_x_log10() +
  facet_wrap(~position) +
  theme_minimal() +
  labs(title="Distance from peak to predicted TSS",
       x="log10(Distance)",
       y="Count")
```








